---
title: "Unstructured"
author: "Homework 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Data

The data is on Canvas in the `data` folder. You will be working with the data in the file called *wweCalls*. Do not unzip it through your computer's GUI; instead, write the code to do it:

In R:

```{r, eval = FALSE}
unzip(zipfile = "path_to_file", 
      exdir = "directory_for_files")
```

Or in Python:

```{python, eval = FALSE}
import zipfile

with zipfile.ZipFile('path_to_file', 'r') as zipper:
  zipper.extractall('directory_for_files')
```

Note that nothing happens in your console when you run these, but the unzipped files will appear where you told them to.

## Step 1

Read all of the *parsed* transcripts into R or Python. You could do them individually, but that is a horrible idea. Instead, use one of the following functions to list files in your directory and read them all in together:

In R:

```{r, eval = FALSE}
list.files("path_to_folder")
```

In Python:

```{python}
import glob

glob.glob('path_to_folder_and_pattern')
```

No matter the language, you will probably want to look at the documentation for the noted functions.

All of these data frames need to be gathered into one data frame. Use some regular expressions to do any text tidy work that you deem necessary.

## Step 2

Register for a free API key from <a href"https://www.alphavantage.co/documentation/">alphavantage</a>. Using your API key, get the daily time series for the given ticker and explore the 10 trading days around each call's date (i.e., the closing price for 5 days before the call, the closing price for the day of the call, and the closing price for the 5 days after the call). If those dates fall on weekends or holidays, do not worry about them. Using the `seq.Date()` in R function will be helpful for generating those date sequences. If you are doing this in Python, I would suggest using `datetime.timedelta` or `datetime` in conjunction with `pd.date_range`. 

## Step 3

Although it is not everyone's cup of tea, professional wrestling (i.e., sports entertainment) is big business -- television contracts, merchandise, and touring generate a lot of money. The site, wrestlingdata.com, has tons of data you never new existed (or wanted to know). One chunk of data is pay-per-view buy rate -- there are around 300 rates across the 6 tables here: https://www.wrestlingdata.com/index.php?befehl=quoten&art=3&liga=1&show=&sort=1&seite=1

You goal is to scrape all 6 pages of buyrate tables, get them into 1 data frame, and plot the absolute buyrate over the dates. What can you infer about wrestling's popularity over the years?